---
title: 'Capstone Project: Billionaires Statistics Analysis'
author: "Nicolas Ojeda"
date: "11/29/23"
output:
  rmarkdown::pdf_document:
    keep_tex: true
    number_sections: true
    toc: true
    toc_depth: 3
    latex_engine: xelatex
  bookdown::pdf_document2:
    keep_tex: true
    number_sections: true
    toc: true
    toc_depth: 3
    latex_engine: lualatex
documentclass: report
papersize: a4
fontsize: 12pt
links-as-notes: true
linkcolor: blue
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
```

\newpage

# Introduction

The distribution of wealth among billionaires is not just a topic of
significant interest in economic circles; its a lens through which we
can view broader economic health, policy effectiveness, and social
equity. This capstone project delves into the Billionaires Statistics
Dataset, a comprehensive aggregation of the worlds wealthiest
individuals, to unravel the complex web of factors contributing to
wealth accumulation among the ultra-rich.

In this analysis, we employ two key machine learning techniques: linear
regression and Random Forest. These methods were chosen for their
robustness and versatility in handling large datasets with multiple
variables. Linear regression will aid in understanding the linear
relationships between various factors and billionaire wealth, while
Random Forest will allow us to explore non-linear patterns and the
relative importance of different predictors.

The overarching goal of this project is twofold: firstly, to predict and
quantify the impact of various factors such as industry, philanthropy,
and geographic distribution on the wealth of billionaires; and secondly,
to identify key drivers and patterns in wealth accumulation that go
beyond mere wealth distribution. Through this analytic journey, we aim
to provide insights that might inform economic policies and contribute
to discussions on wealth inequality and its broader implications.

\newpage

# Literature Review

This project, which centers on the analysis of billionaire wealth using
linear regression and Random Forest models, aligns seamlessly with
recent advancements in predictive analytics and the widespread
application of machine learning techniques across diverse fields.

In the realm of predictive analysis, a study titled "An Overview of the
Supervised Machine Learning Algorithms" offers a comprehensive
comparison between the Generalized Linear Model with Linear Regression
(LR) and Decision Trees with Random Forest (RF). This research
underscores the efficacy of both models in handling various types of
data. While linear regression demonstrated superiority in a specific
analysis scenario, the study emphasizes the potential of combining LR
and RF predictions. Such an approach not only yields more accurate
results but also mitigates the risks associated with overfitting and
incorrect predictions that may arise when relying solely on individual
algorithms. These insights into the synergy between linear regression
and Random Forest models are highly relevant to our project's
objectives.

Furthermore, the practical application of these models in precision
agriculture is exemplified in the study titled "Using Linear Regression,
Random Forests, and Support Vector Machine with Unmanned Aerial Vehicle
Multispectral Images to Predict Canopy Nitrogen Weight in Corn." This
research leveraged UAV multispectral imagery to predict canopy nitrogen
weight in corn fields. The study's findings highlighted the superior
performance of Random Forests over linear regression, emphasizing the
model's robustness in handling non-linear data and its ability to
provide accurate predictions even when dealing with low variance data.
Notably, the study employed the R programming language and the Random
Forests algorithm, underscoring the utility of these models in handling
complex datasets with multiple variables.

Collectively, these studies underscore the versatility and effectiveness
of linear regression and Random Forest models in predictive analytics.
Their applications span from software cost estimation to precision
agriculture, illustrating the broad spectrum of utility across diverse
domains. In our project, we draw upon these insights to analyze
billionaire wealth, harnessing the power of both linear regression and
Random Forest models to uncover patterns and predict influential factors
in wealth accumulation.

\newpage

# Methodology

Data Collection: The primary data set for this project, the Billionaires
Statistics Data set, was sourced from
<https://www.kaggle.com/datasets/nelgiriyewithana/billionaires-statistics-dataset>
. This data set offers a comprehensive aggregation of financial and
personal details of the worlds wealthiest individuals, including their
net worth, sources of wealth, involvement in various industries,
philanthropic activities, and more. This data set provides a robust
foundation for analyzing the factors influencing billionaire wealth.

# Data Wrangling

The data set underwent in 2 preprocessing steps to ensure its quality
and suitability for analysis. These steps included:

Data Cleaning:Identification and imputation of missing values were
carried out to ensure data completeness. The data was also standardized
for consistency in formatting, and duplicate entries were removed.

Outlier Detection Outlines: were identified using statistical methods
and addressed to prevent skewing of the analysis.

\newpage

# Exploratory Data Analysis

# Data Preparation

In this section, we install, and load every packages required for this
project,

```         
```

```{r}
# Load all packages required for the project
required_packages <- c(
  "tidyverse",    
  "lubridate",    
  "ggplot2",     
  "dplyr",        
  "readr",       
  "caret",        
  "randomForest",  
  "rmarkdown", 
  "stats"
)

# Install missing packages
new_packages <- required_packages[!required_packages %in% installed.packages()[,"Package"]]
if(length(new_packages)) install.packages(new_packages)

# Load all required libraries
invisible(lapply(required_packages, library, character.only = TRUE))

# Load the dataset (assumes CSV format)
billionaires <- read.csv("C:/Users/nico0/OneDrive/Documents/billionare/Billionaires Statistics Dataset.csv", stringsAsFactors = FALSE)

# Preliminary data cleaning
billionaires <- billionaires %>%
  mutate_if(is.character, trimws) %>%
  na.omit()

# Summary statistics
summary(billionaires)

```

# Total Billionaire Wealth by Country

```{r}
ggplot(billionaires, aes(x = country, y = finalWorth)) +
  geom_col(fill = "blue", color = "black") +
  labs(title = "Total Billionaire Wealth by Country", 
       x = "Country", 
       y = "Total Wealth (in billions)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

This bar chart shows the sum of wealth for each country. There is a
significant variance in total wealth between countries. The United
States stands out with the highest total wealth, which could suggest it
has either a greater number of billionaires or higher individual wealth
values, or both. The distribution is highly skewed, with most countries
having considerably less total wealth compared to the United States.

# Wealth Distribution by Category

```{r}
ggplot(billionaires, aes(x = category, y = finalWorth)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Wealth Distribution by Category", 
       x = "Category", 
       y = "Wealth (in billions)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This dot plot presents the wealth distribution across different industry
categories. It seems to be displaying individual data points for wealth
within each category, which could represent individual billionaires
wealth in each sector. Some categories like Technology, Finance &
Investments, and Manufacturing appear to have a wider range of wealth
values, with some individuals significantly wealthier than others within
the same category. There may be outliers in several categories that
could represent particularly successful individuals or industry giants.

### **Distribution of Billionaire Ages**

```{r}
# Assuming your dataset has a column named 'age'
hist(billionaires$age, 
     breaks = 30, 
     main = "Distribution of Billionaire Ages", 
     xlab = "Age", 
     ylab = "Frequency", 
     col = "lightblue", 
     border = "black")

```

This histogram displays the frequency distribution of ages among
billionaires. The x-axis represents different age groups, while the
y-axis shows the frequency of billionaires within those age groups. The
distribution appears to be right-skewed, indicating that there are fewer
young billionaires and a greater number of older billionaires. The
majority of billionaires fall within the middle age brackets, which
could suggest that wealth accumulation peaks during these years. The
skewness towards older ages may reflect the time it takes to build and
amass significant wealth.

# **Cumulative Wealth by Age**

```{r}
ggplot(billionaires, aes(x = age, y = finalWorth)) +
  stat_ecdf(geom = "step", color = "darkblue") +
  labs(title = "Cumulative Wealth Distribution by Age", 
       x = "Age", 
       y = "Cumulative Wealth (in billions)") +
  theme_minimal()

```

Represents a step graph that shows the cumulative distribution function
of wealth as it relates to the age of billionaires. The x-axis displays
the age, while the y-axis shows the cumulative wealth. The graph
indicates that wealth accumulation increases with age, suggesting that
the older billionaires tend to have higher wealth accumulation, which
could be due to more extended periods of wealth generation and
compounding investments over time.

# **Cumulative Wealth by Age**

```{r}
ggplot(billionaires, aes(x = finalWorth)) +
  geom_density(fill = "salmon", color = "darkred", alpha = 0.7) +
  labs(title = "Density Plot of Billionaires' Wealth", 
       x = "Wealth (in billions)", 
       y = "Density") +
  theme_bw() +
  xlim(0, quantile(billionaires$finalWorth, 0.99)) # Limiting x to the 99th percentile for better visualization

```

This s a density plot that provides a smoothed representation of the
distribution of billionaires' wealth. The x-axis represents the wealth
in billions, and the y-axis represents the density of the probability
distribution for wealth. The plot shows a peak at the lower end of the
wealth spectrum, indicating a high density of billionaires with
relatively lower wealth, and a long tail extending towards the higher
wealth values, representing the rarity of extremely high wealth. This
pattern reflects the inequality in wealth distribution, with a large
number of billionaires having wealth in the lower range of the spectrum
and a few individuals having significantly higher wealth.

\newpage

# Model Selection

# Linear Regression

What is Linear Regression? Linear Regression is a statistical method
used to model the relationship between a dependent variable and one or
more independent variables. In this analysis, it is used to understand
how various factors, such as age, industry, country, and self-made
status, linearly correlate with the wealth of billionaires.

# Model Development

# Linear Regression Analysis for Gender Differences

```{r}
# Splitting the data by gender
data_male <- filter(billionaires, gender == "M")
data_female <- filter(billionaires, gender == "F")

# Inspecting all factor variables in the male dataset
sapply(data_male, function(x) if(is.factor(x)) levels(x))

# Inspecting all factor variables in the female dataset
sapply(data_female, function(x) if(is.factor(x)) levels(x))

# After checking, adjust your model accordingly
# Build the linear regression model for the male and female datasets
# You might need to exclude factor variables with only one level or convert them to numeric/dummy variables

# Example model (adjust based on your findings and the unique values in each subset)
model_male <- lm(finalWorth ~ age + country + source, data = data_male)
model_female <- lm(finalWorth ~ age + country + source, data = data_female)

# Summary of the models
summary(model_male)
summary(model_female)

```

# Results

# Lineal Regression

```{r}

summary(model_female)
summary(model_male)


```

# Data Description

The dataset used in this analysis consists of individuals' information,
including their age, country of residence, and the source of their
wealth. Two separate models were developed, one for males and one for
females.

#Model Specification The linear regression models were specified as
follows:

For Males and females

#finalWorth \~ age + country + source

# Results and Interpretation

# Analysis for Females

Coefficients

The intercept (-1725.63) represents the estimated "finalWorth" when all
predictor variables are zero. The "age" coefficient (42.41) suggests
that for each additional year of age, "finalWorth" is expected to
increase by approximately 42.41 units. Several "country" coefficients
indicate the impact of different countries on "finalWorth."
Significance: The level of significance for each coefficient varies.
Notably, "country10" is statistically significant at the 0.05
significance level.

Residual Standard Error

The residual standard error (8233) measures the spread of errors around
the regression line. R-squared: The multiple R-squared value (0.0504)
suggests that only about 5.04% of the variance in "finalWorth" is
explained by the model.

Adjusted R-squared

The adjusted R-squared is negative (-0.01434), indicating that the model
may not be an ideal fit for the data. F-statistic and p-value .The
F-statistic (0.7785) and associated p-value (0.7249) suggest that the
model may not be statistically significant.

# Analysis for Males

Coefficients

The intercept (676.40) represents the estimated "finalWorth" when all
predictor variables are zero. The "age" coefficient (38.53) suggests
that for each additional year of age, "finalWorth" is expected to
increase by approximately 38.53 units. Various "country" and "source"
coefficients indicate the impact of different countries and sources on
"finalWorth." Significance

Some coefficients, such as "country10" and "age," are statistically
significant at the 0.05 significance level. Residual Standard Error:

The residual standard error (10440) measures the spread of errors around
the regression line. R-squared

The multiple R-squared value (0.01615) suggests that only about 1.62% of
the variance in "finalWorth" is explained by the model. Adjusted
R-squared

The adjusted R-squared (0.006273) remains low, indicating limited model
fit. F-statistic and p-value

The F-statistic (1.635) and associated p-value (0.03437) indicate some
level of statistical significance, though the model's explanatory power
is limited.

# Random Forest

What is Random Forest? Random Forest is an ensemble learning method
known for its robustness and versatility. It operates by constructing a
multitude of decision trees during training and outputs the class that
is the mode of the classes (classification) or mean prediction
(regression) of the individual trees.

# Random Forest model for predicting wealth of billionaries

```{r}

# Converting categorical variables to factors
billionaires$country <- as.factor(billionaires$country)
billionaires$source <- as.factor(billionaires$source)
billionaires$category <- as.factor(billionaires$category)

# Convert 'age' to numeric if it's not already
billionaires$age <- as.numeric(billionaires$age)

# Optional: Remove rows with NA values if they exist
billionaires <- na.omit(billionaires)

# Finding top 10 countries
top_countries <- names(sort(table(billionaires$country), decreasing = TRUE)[1:10])

# Finding top 10 sources
top_sources <- names(sort(table(billionaires$source), decreasing = TRUE)[1:10])

# Reducing the number of levels in categorical variables
# Group countries into fewer categories
billionaires$country <- as.factor(ifelse(billionaires$country %in% top_countries, billionaires$country, "Other"))

# Repeat similar steps for 'source' if it has more than 53 levels
# Replace 'top_sources' with a vector of your selected sources
billionaires$source <- as.factor(ifelse(billionaires$source %in% top_sources, billionaires$source, "Other"))

# Building the Random Forest model
model_rf <- randomForest(category ~ age + country + source, data = billionaires, ntree = 100)

# Viewing the model summary
print(model_rf)


```

# Results

# Random Forest Model Outcomes

```{r}
importance(model_rf)




```

# Data Preprocessing 

To prepare the data for modeling, several preprocessing steps were
undertaken:

Categorical Variables Categorical variables, including "country,"
"source," and "category," were converted into factors. This conversion
allows us to use them as predictors in our model effectively.

Numeric Conversion:The "age" variable was converted into a numeric
format to facilitate modeling.

Handling Missing Data Rows with missing values were removed from the
dataset to ensure a clean and complete dataset for modeling.

Exploratory Analysis To gain insights into the data, an exploratory
analysis was conducted: The top 10 countries with the highest number of
billionaires were identified. Similarly, the top 10 sources of wealth
were identified.

Feature Engineering To simplify the modeling process and improve model
generalization, the following feature engineering decisions were made:

Country Grouping: Countries were grouped into fewer categories. The top
countries were retained, while the rest were labeled as "Other." This
simplifies the model without losing essential information. Building the
Random Forest Model The next step involved building the Random Forest
model:

The Random Forest model was constructed using the "category" variable as
the target variable and "age," "country," and "source" as predictor
variables.

A total of 100 trees were used in the Random Forest model.

Model Summary The model summary provides an overview of the Random
Forest model: It specifies the model type, the number of trees used
(100), and the number of variables considered at each split in the
trees.

Additionally, it includes an out-of-bag (OOB) estimate of the error
rate, which indicates the model's accuracy in predicting wealth
categories. Model Outcomes The importance of predictor variables in the
Random Forest model was evaluated using the Mean Decrease in Gini index:
The Mean Decrease in Gini measures the importance of each variable in
making accurate predictions. Higher values indicate greater importance
in the model.

# Conclusion

In this analysis, I leveraged two machine learning models, Random Forest
and Linear Regression, to delve into the wealth and financial status of
billionaires. My objective was to uncover insights into the predictive
power of variables like age, country of residence, and source of wealth.

# Random Forest Model

In developing the Random Forest model, I embarked on a journey to
predict the wealth category of billionaires. Here are the key takeaways
from my exploration of the Random Forest model the Random Forest model,
composed of 100 trees, was meticulously trained to forecast the wealth
category of billionaires. I relied on critical features such as age,
country, and source of wealth.

The model's out-of-bag (OOB) estimate of the error rate, standing at
approximately 60.03%, provided insights into its predictive accuracy. It
appears that the model may have limited precision when categorizing
billionaires' wealth based on the chosen features.

To assess the importance of predictor variables within the model, I
employed the Mean Decrease in Gini index. My analysis unveiled that age,
country, and source of wealth played pivotal roles in facilitating
accurate predictions.

For further model enhancement, I acknowledged the necessity of refining
the model and engaging in strategic feature engineering. Future steps
may involve exploring alternative machine learning algorithms and
advanced feature selection techniques to bolster predictive
capabilities.

# Linear Regression Model

In tandem with the Random Forest model, I embarked on a journey with the
Linear Regression model to elucidate the intricate relationships between
age, country, source of wealth, and the financial worth of billionaires.
Here are my key findings and inferences:

I painstakingly constructed the Linear Regression model to predict the
financial worth of billionaires (finalWorth) predicated on age, country,
and source of wealth.

Through the model's coefficients, I shed light on the complex interplay
between predictor variables and the financial worth of billionaires.
Notably, age emerged as a significant positive predictor, suggesting
that as age increases, financial worth tends to rise.

The adjusted R-squared value, while informative, underscored that the
selected predictor variables only partially elucidate the variance in
financial worth. The model, as it currently stands, does not encapsulate
the entire spectrum of wealth accumulation dynamics.

Importantly, the Linear Regression model demonstrated moderate
predictive capability. The path forward involves exploring additional
variables or adopting more intricate modeling techniques to enhance
predictive accuracy.

In summary, my dual-pronged approach, encompassing both the Random
Forest and Linear Regression models, unearthed crucial insights into the
factors influencing billionaires' financial success. While the Random
Forest model revealed limitations in wealth category prediction, the
Linear Regression model provided a nuanced yet incomplete understanding
of the factors underpinning billionaires' financial prosperity.

For future endeavors, I am inclined to focus on augmenting the
predictive efficacy of the Random Forest model and delving deeper into
the determinants of wealth. This may entail incorporating additional
variables or harnessing more sophisticated modeling methodologies to
enrich our understanding of billionaires' financial trajectories.

# References

<https://www.intechopen.com/chapters/84394>

<https://www.mdpi.com/2072-4292/12/13/2071>
